---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?
**A1.** 26 departments, Atospheric and Oceanic Science offered the most classes with "Climate" in the title

```{r}
and_department_seats_1 <- umd_courses |>
  filter(str_detect(title, "Climate") & seats >= 1)|>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats)
  ) |>
  arrange(desc(total_seats))
```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why?
**A2.** I guess I'm interested in the department called "Honors" because I don't know what that is. I'm also interested to know what meaning of "climate" each title is using. It could be the scientific meaning or the general meaning, like "a climate of fear." If for example the criminology class is about the scientific meaning of climate I'd be interested to find out more about those classes and how those topics intersect.


**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?

**A3.** Business and Management has the most courses in the Fall 2023 term. 

Business and Management: 164 classes
School of Music: 137 classes
English: 120 classes
Teaching and Learning, Policy and Leadership: 116 classes
Information Studies: 115 classes

Top 5 departments with least one syllabus posted:
Business and Management: 82 classes
Economics: 50 classes
School of Music: 48 classes
Biological Sciences Program: 41 classes
Information Studies: 41 classes

125 departments have at least one syllabus posted in that term. 64 departments in that term don't have any syllabi posted.

Business and Management and School of Music are in the top 5 of both lists, the other 3 results are different.

```{r}
departments_202308 <- umd_courses |> filter(term == 202308)
departments_202308 |> 
  filter(syllabus_count>=1)|>
  group_by(department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))

```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.
**A4.** The data shows the monthly expenses of Moore Wes For Maryland from May 2021 to November 2022. The first month's expenses are 6,000.3 which might be spending money from one maximum donation to the campaign. The next month spending jumps to 21,000. Then spending roughly ramps up over time, with some variation, and peaks the month before the election, which makes sense. 

```{r}
maryland_expenses |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  filter(str_detect(committee_name, "Moore  Wes For Maryland")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount)) |>
  arrange(month)
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5** I created a new dataframe called in_state_out that shows how much money was spent in state and out of state. It shows 77,723,146 was spent in state and says NA for out of state. I don't know why it says NA because the maryland_expenses dataframe shows out of state expenses.

```{r}
in_state_out <- maryland_expenses |>
  mutate(
    spending_location = case_when(
        str_detect(address, " Maryland ") ~ "In-state",
        .default = "Out of state"
    )
      )|>
group_by(spending_location) |>
  summarize(total = sum(amount)) |>
  arrange(desc(total))
```


**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.
**A6.** 

```{r}
```
