---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.

```{r}
library(tidyverse)
library(lubridate)
library(tidycensus)
```

## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

* Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.
```{r}
prince_georges_2022_overdoses <- read_csv("data/prince_georges_2022_overdoses.csv")

pg_2022_overdoses_date_week_month <- prince_georges_2022_overdoses |>
  
  mutate(
    month = month(datetime, label=TRUE),
    week = week(datetime),
    date = date(datetime)
    )

```



## Questions 

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1.** (Worked alongside Ilana, Mekhi and April)

We saw there were 1397 rows in the original dataframe which represents 1397 total calls.

The month that had the greatest percentage of total overdose calls is December: 141 calls/10.1% of all calls.

I couldn't see a clear pattern in the bar graph organized by month so I tried organizing it by week. I saw a big spike in one week, but it was still not super clear what the overall trends were. I looked at the week dataframe and saw the weeks with the most calls were week 28, 51, and 48. In 2022: Week 28 was July 11-17, week 51 was Dec 19-25 and week 48 was Nov 28-Dec 4. Though those dates might be a bit off because the original data seemed to count weeks strangely: they had 53 weeks in 2022.

It did look like there are spikes around Christmas, Thanksgiving and 4th of July.

To compare the results to the Baltimore County data:

Baltimore County had more total calls overall. The calls also seemed to increase in warmer months and December was the second lowest in total calls. In PG County December was highest for total calls. In Baltimore County August was highest for total calls but it was second lowest in PG County.

```{r}
pg_2022_overdoses_by_month <- pg_2022_overdoses_date_week_month |>

  group_by(month) |>
  summarise(total_calls = n()) |>
  mutate(pct_calls = (total_calls/1397)*100) |>
  arrange(desc(pct_calls))

pg_2022_overdoses_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,pct_calls), weight=pct_calls)) +
  geom_bar(aes(x=month, weight=pct_calls)) +
  labs(
    title="More 911 Overdoses Around National Holidays in PG County",
    x = "month",
    y = "percent of total 911 overdose calls",
    caption = "source: Prince George's County EMS"
    
  )

pg_2022_overdoses_by_week <- pg_2022_overdoses_date_week_month |>

  group_by(week) |>
  summarise(total_calls = n()) |>
  mutate(pct_calls = (total_calls/1397)*100) |>
  arrange(desc(pct_calls))

pg_2022_overdoses_by_week |>
  ggplot() +

  geom_bar(aes(x=week, weight=pct_calls)) +
  labs(
    title="More 911 Overdoses Around National Holidays in PG County",
    x = "week",
    y = "percent of total 911 overdose calls",
    caption = "source: Prince George's County EMS"
    
  ) +
  theme(
    axis.text.x = element_text(angle = 90,  hjust=1)
  )

```

**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2.** The plot shows the average range of number of calls was between 20 and 30. There's a big spike in July and increases in November and December and then a big drop off in January. (See earlier answer for analysis of the three weeks with the most calls.)


```{r}

#already made a data frame by week

pg_2022_overdoses_by_week |>
  ggplot() + 
  geom_line(aes(x=week, y=total_calls)) +
  labs(
    title="Prince George's Overdoses Spike in July, Plummet in January",
    x = "week",
    y = "991 overdose calls",
    caption = "source: Prince George's County EMS"
    
  )
```

**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.
2. Get data from the Census Bureau showing median household income for Maryland zip codes.
3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5. Give it an appropriate title, source, and x and y axis titles.
6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.
7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3.**  I wanted it to be "geom_scat" but April and I found using Chat GPT that it is in fact "geom_point."
Paul helped me take the info from your geom_text doc and implement it. I messed with the hjust to make sure the labels don't overlap the points. I tilted the labels a little because one on the right was getting cut off by the edge of the plot. I can't use check overlap because it gets rid of some of the labels.

I decided to put median income on the the x-axis and total calls on the y-axis because the thesis is about how many calls there are and I associate "up" with "more" so I think it's easier to read that way. 

7. Yes, I think median household income does have a relationship with overdose calls in that zip code. I see the biggest chunk of calls happen in zip codes with median household income that's around 75,000 dollars a year. That's on the low end of middle class income in Prince George's County, on the edge of qualifying for housing assistance if those households have incomes from 1 or 2 people. Some of the lowest income zip codes don't have high numbers of calls which may be because it's easier to pay for a substance abuse problem if you have some disposable income? I think that makes sense because there are quite a few calls in zip codes with incomes a bit more than 100,000 dollars.

I think I'd report the two spikes in calls and how at very low and very high income the number of calls drops off.


```{r}

pg_2022_overdoses_zip <- prince_georges_2022_overdoses|>
  group_by(zipcode) |>
  summarise(total_calls = n()) |>
  arrange(desc(total_calls))

  md_zip_medincome <- get_acs(geography = "zcta",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2019)
  
pg_2022_overdoses_zip_income <-pg_2022_overdoses_zip |> left_join(md_zip_medincome, by= c('zipcode'='GEOID')) |>
  rename(med_income=estimate)


 p <- ggplot(data = pg_2022_overdoses_zip_income, aes(x=med_income, y=total_calls)) +
  geom_point(shape = 19, color = "red", size = 2.5) +
  labs(
    title="Lower-Income PG County Zip Codes Have More Overdoses",
    x = "Median houshold income",
    y = "Total 911 overdose calls",
    caption = "Source: Prince George's County EMS and 2019 Census"
    
  )


p + geom_text(aes(label = zipcode),
  hjust = -.25, 
  nudge_x = 0.05, 
  size = 2.5, 
  angle = 20,
  check_overlap = FALSE
)
  
```
