---
title: "Grad Student Assignment"
author: "Cait Kelley"
date: "12/3/2023"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this notebook, I'm looking at violent crime and property crime stats in Maryland: <https://opendata.maryland.gov/Public-Safety/Violent-Crime-Property-Crime-by-County-1975-to-Pre/jwfa-fdxs>

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(tidycensus)
library(janitor)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data.

Other questions to answer in this notebook:

Who created and maintains the data?

What state program or law is connected to it?

What does each row represent?

What is the chronological and geographic scope of this data?

If this data contains aggregates (totals), can you find itemized data that those totals are derived from?

Do the totals match your own calculations using the itemized data?

```{r}
# Load required data
violent_and_property_crime_md <- read_csv("data/violent_crime_property_crime_by_county_1975_to_present.csv")


# Clean required data and prepare for analysis if needed. 


```

## Story Ideas

In this notebook, I am exploring three potential story ideas.

### Story Idea 1

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Story Idea 2

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Story Idea 3

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

