---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Look at the earthquakes dataset. Finish the sentence below as if you were explaining the data to someone who had not seen it before but needs to know about it.

**A1:** This dataset contains data about all earthquakes that occured in the world between September 11, 2023 and October 11, 2023. 

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2:** There are 9774 records. The dataset includes useful information like the size of the earthquakes, their location, type of earthquake, and date and time.

------------------------------------------------------------------------

#### **Q3** How do I reorganize this data to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes<-earthquakes|>
 
  arrange(desc(depth))
```

**A3:** The deepest earthquake was 669.982 with magnitude 4.2.

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
earthquakes|> filter(mag>6)
```

**A4:** 13 

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
earthquakes|>filter((mag>6)&(depth<20))


```

**A5:** There are 6 earthquakes with mag larger than 6 and depth smaller than 20.

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 OR a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
earthquakes|>filter((mag>6)|(depth<20))
```

**A6:** 7,446 earthquakes in the dataset had a mag greater than 6 or a depth less than 20.

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
earthquakes|>
  filter(str_detect(place, "Alaska"))

```

**A7:** Look at the "place" column and string detect for "Alaska." Not all the net id's for those locations say "ak," some say "us," so don't string detect based on the net column. 3,446 earthquakes happened in Alaska.

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
earthquakes|>
  group_by(type)|>
  summarise(
    count_type = n()
  )

```

**A8:** The first two most common types of tremors are earthquakes and explosions.

------------------------------------------------------------------------

#### **Q9** What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
earthquakes|>
  filter(type=="earthquake")|>
  summarise(
    mean_depth = mean(depth)
  )
  
```

**A9:** Average depth of earthquakes in the dataset: 25.23796 kilometers
The average depth seems very shallow to me which is interesting. The shallower the earthquake, the more destructive it is.

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
earthquakes|>
  mutate(hour= hour(time))|>
arrange(desc(hour))
```

**A10:** Derek, in the future you can mutate the data to extract the hour. That will make a new column displaying just the hour of each tremor.

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column showing whether an earthquake is serious or not?

```{r}
earthquakes |>
  mutate(severity = case_when(
       mag>3 ~ "serious",
        .default = "not serious")
    )

```

**A11:** You use mutate and case when to say all earthquakes with magnitude bigger than 3 are "serious" in a new column called "severity" and all other earthquakes are "not serious."

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
earthquakes |>
  mutate(hour= hour(time))|>
  mutate(severity = case_when(
       mag>3 ~ "serious",
        .default = "not serious"))|>
         group_by(hour, severity)|>
  summarise(
    count_per_hour = n())|>
arrange(desc(count_per_hour))
  
```

**A12**: 12:00 pm has the fewest serious earthquakes (33). Arange by descending and then click to next pages to see hour with most serious eathquakes, arrange by ascending to see hour of fewest serious earthquakes. 2:00 am has the most serious earthquakes (67).

------------------------------------------------------------------------

#### **Q13** What's another question you are interested in and how would you ask it in R?

What time of day has the most not serious earthquakes? A: 11:00 am has 397 tremors.

```{r}

earthquakes |>
  mutate(hour= hour(time))|>
  mutate(severity = case_when(
       mag>3 ~ "serious",
        .default = "not serious"))|>
         group_by(hour, severity)|>
  summarise(
    count_per_hour = n())|>
arrange((count_per_hour))



```
